Some thought experiments on the applicability of Granger causality and Directed Information in inferring direction of information flows

### 2015-07-06 paper.tex
Edit introduction. Add references.

This work is in large part motivated by a recent surge of interest in understanding neural circuits --- the connectivity and dynamic activity of parts of the brain --- and how they give rise to behavior and experience. This is evidenced in the launching of the BRAIN initiative in the US and the Human Brain Project in Europe. To quote from \emph{BRAIN 2025: A Scientific Vision}\footnote{The BRAIN Working Group's report to the Advisory Committee to the Director of the NIH}, we wish to ``map connected neurons in local circuits and distributed brain systems, enabling an understanding of the relationship between neuronal structure and function'', clearly indicating the move towards understanding (a) the connectivity and (b) the computational function of different brain regions. While the question of how the brain computes has been of immense interest for several decades, only recently have measurement techniques become sophisticated enough to be able to simultaneously record the activity of multiple neurons, or multiple neural populations.

In an effort to make headway on these goals, several works use Granger causality (and less often, its information-theoretic generalization -- Directed Information) to understand how information flows in the brain ({\color{red} ??}), or to acquire directed maps of functional connectivity ({\color{red} ??}). For instance, in~\cite{Brovelli2004BetaOscillations}, Granger causal influences that are measured between somatosensory and motor sites are said to ``support the idea that somatosensory feedback provides information to the sensorimotor system that is used to control motor output''. This raises the question: do these directed connectivity maps, as determined by directional causal influence measures such as Granger causality, correctly identify the directions along which information flows in the brain?

\subsection{How Granger Causality is used in Neuroscience today}

We take~\cite{Brovelli2004BetaOscillations} and~\cite{Ding1999ShortWindow} to be the canonical references for the application of Granger causality to estimate causal influences, as they seem to be widely accepted by the community (\cite{Bressler2011WienerGranger},{\color{red} \cite{Roebroeck2005MappingDirected}},~\cite{Ding2006GrangerCausality}, \cite{Barnett2014MVGC}). Here, we briefly describe how Granger Causal influence is quantified, and how it is computed in these works.

Granger causality, as originally described by Granger~\cite{Granger1969InvestigatingCausal}, measures the level of influence that one process $\{X\}$ has on another process $\{Y\}$. The analysis compares the \textit{error in predicting} the $\{Y\}$ process based on (i) simply the past of $\{Y\}$, and (ii) based on the past of both $\{X\}$ and $\{Y\}$. The Granger causality metric is the ratio of these errors. Many variations of Granger causality have also been considered, including a generalization --- the Directed Information (see~\cite{MasseyCausality,quinn,weissman}) --- an information-theoretic quantity denoted by $I(\m{X}\to\m{Y})$.

Granger causality computation, in general, requires that the processes in consideration be stationary. In neuroscientific experiments, this is usually not the case, as the express objective of the experiment is to understand brain function while the brain is subjected to a task. In order to circumvent this problem, data analysis techniques either assume stationarity~\cite{Brovelli2004BetaOscillations}, or work with time-windows short enough that the processes are in fact stationary within those windows~\cite{Ding2006GrangerCausality}. In the latter case, Granger causal influences are computed for each window individually, in order to produce a dynamic description of brain connectivity as the brain carries out the said task.

In order to determine the direction of causal influence, the Granger Causality metric (ratio of residual variances) from $\{X\}$ to $\{Y\}$ is \emph{compared} to that from $\{Y\}$ to $\{X\}$. We note here that Granger's original analysis does not compare this metric on forward and reverse links, however, this appears to be commonly done in practice. {\color{red} TODO: Cite Tsachy's paper here.}

@@

\subsection{A short survey of previous criticisms of Granger Causality}

@@

\begin{enumerate}
	\item First, Granger Causality suffers from what we call the ``hidden node problem''. If two observed nodes receive causal influences from a third, latent node, then a Causal influence may be detected between the observed nodes, even if they are independent of each other~\cite{PearlCausalitychapterasdfads}. All nodes need to be observed, therefore, to avoid finding spurious influences.
	\item Second, if the measurements from each node are differentially affected by noise, then the predicted direction of causal influence might be opposite to the true direction of influence~\cite{Friston34234, EcoPaper3q4234}. Measurements need to be relatively noiseless and precise in order to obtain the correct direction of causal influece.
	\item Third, subsampling the processes can produce incorrect results~\cite{KunZhangsPaper}. {\color{red} TODO: Write another line here - look at the paper for this.}
\end{enumerate}

It is important to note that the technical objections listed above are all concerned with measurement. They indicate that incorrected Granger Causal influences may be computed if there is some deficiency in the measurement process. These objections can be resolved by taking better measurements (by sampling more nodes, with higher signal-to-noise ratio, or {\color{red} by not subsampling}).

Our objection, on the other hand, is much more fundamental. \emph{Even if} the measurements are made with infinite accuracy, and the regression coefficients associated with computing Granger Causality are precisely estimated, and the Granger Causality metric is perfectly computed (as is the case in our counter-examples), Granger Causal influence would \emph{still} not give the correct direction of information flow. To our knowledge, this is a completely novel argument against Granger Causality, and one that is far more serious than previous objections, at least as far as its use in neuroscience is concerned.

\subsection{Our counter-examples and our main result}

{\color{red} Needs re-write}

This paper considers two experiments (introduced in Section~\ref{sec:sk}) where a transmitter $Tx$ wants to communicate a message to a receiver $Rx$ in presence of a feedback channel (in one experiment, the feedback link is noiseless, while in the other it is noisy). We assume that the experimenter is able to record the transmissions of $Tx$ and $Rx$ using some probing mechanism. Provided with these measurements, the experimenter wants to estimate which node is $Tx$ and which is $Rx$. Our results, derived in Section~\ref{sec:analytical-results} and numerically illustrated in Section~\ref{sec:numerical-results}, show that both Granger causality and Directed Information can incorrectly estimate which node is which. The first experiment is for the (unrealistic) case of communicating across noiseless feedback channels. The second experiment allows for noise in the feedback channel as well. In both cases, linear strategies inspired from the strategy of Schalkwijk and Kailath~\cite{S&K} are used.

Our goal here is to bring out the point that whether Granger causality and Directed Information can be used to interpret direction of information flows is an issue that can be, and perhaps should be, considered using thought experiments on simple communication problems where \emph{information flow direction, and quantity, is already known}. If the prediction of Granger causality, or some other measure of information flow, matches with the known information flow, then that measure can be more confidently used in experiments. While our results strongly suggest that one needs to exercise care in interpreting Granger causality and Directed Information dominance as an indicator for the direction of information flow, there are several shortcomings that need to be addressed in order to understand the issue at depth. These shortcomings are discussed in detail in Section~\ref{sec:conclusions}.

\begin{enumerate}
	\item Define Granger causal influence and directed info in English again. Define information flow (use examples, or define formally? Formal definition should come in the math sections). Describe the problem again - is the direction of causal influence the same as the direction of information flow?
	\item Describe the counter-examples in English.
		\begin{enumerate}
			\item Transmitter has the message (msg is known at the transmitter before it is known at the receiver - this is what distinguishes Tx from Rx). Therefore, the true direction in which information is flowing is *known* to us, by design of the experiment. This is the *ground truth*, against which we can compare the results of Granger causality.
			\item Feedback link is present - we consider two cases: feedback link noiseless, and feedback link noisy.
			\item We consider the case where Tx and Rx are using the S\&K strategy to communicate.
			\item We record time series of transmissions at Tx and Rx, and using this we use Granger Causality to try to understand which direction the information is flowing.
		\end{enumerate}
	\item We discover that for reasonable choices of parameter values, the direction of Granger causal influence is opposite to the direction of information flow.
\end{enumerate}

\subsection{Possible objections to, and shortcomings of this work}

A review of a previous submission raised some objections to this work, which we seek to rectify here. Further, our analysis has certain shortcomings, which we acknowledge here, and could be room for future work.

\begin{enumerate}
	\item Firstly, and most importantly, previous work has already shown that Granger Causality is not true causality (as defined by Judea Pearl~\cite{PearlCausalityBook}. Our argument is, in spirit, very similar to this common criticism of Granger causality, which states that Granger Causal influence is a statistical quantity and that it does not answer the question of true causal influence: if X is turned off, will Y behave differenty? {\color{red} (Need to quote from Pearl here)}.However, in the situation we are studying, we ask a different question - we wish to understand whether or not Granger Causality gives the correct direction of \emph{information flow}. In our analysis, causal influences are present in both directions, even in a ``true-causality'' sense.
	\item It might appear that we make a circular argument while computing the Granger Causal influences in our counter-example, since we supply the model for the stochastic process and use the theoretical coeffiecients of the model directly to compute the Granger Causality metric. However, as we state in Section~\ref{sec:figure-it-out}, we assume that the regression coefficients of the Autoregressive model can be exactly estimated (even if the AR process is non-stationary), and discuss how this might be achieved with the help of multiple trials. {\color{red} (Move to conclusions?)}
	\item Our analysis does not tackle an information source that evolves with time. Hence, our communication process (inspired from the scheme of Schalkwijk and Kailath~\cite{S&K}) is non-ergodic. Since Granger causality is really just relative errors in prediction of a process, in presence or absence of knowledge of another process, we compute the obvious generalization of Granger causality to non-ergodic processes. Nevertheless, future work will address a situation with a linear dynamical system as the information source.
	\item Our experiments restrict themselves to Gaussian noise for simplicity, but neural spiking and spike-rate models for spikes tend to be very different from those used here. This is a clear direction for future work.
	\item The power and energy constraints are somewhat oversimplified to make the analysis simpler. This is for simplicity of exposition. A more general analysis is a trivial extension.
	\item We consider a simple point-to-point network. In general networks, this issue could be even more complex. However, since this issue shows up even for the simplest network, we feel that the problem will only be exacerbated when the network is large.
	\item We would also like to note that this paper is \emph{not} mathematically deep. We believe that this is a strength of our analysis. Even simple counter-examples that do not employ difficult proof techniques are able to show that Granger Causality does not always correctly estimate the direction of information flow. {\color{red} (Move this?)}
\end{enumerate}

\subsection{Discussion}
\begin{enumerate}
	\item Some of this misconception might have arisen because of Massey's directed info conservation - we expect that dir info in the forward dirn goes up if the other one goes down - but this is a misnomer. (Do we still need this? I am increasingly of the opinion that directed info / GC capture nothing about the message)
\end{enumerate}

@@

In order to arrive at such an understanding of the brain, a natural method is to probe it to understand its functional connectivity\footnote{This is different from structural or anatomical connectivity in that links may sometimes be present, but not active.}. Analyses are then performed on the time series thus obtained from the probes, in order to say something about the functional connectivity --- which components influence each other, and which direction is this influence in. One commonly used analysis is based on the computation of Granger causality~\cite{Granger} with the measurements from each node. The Granger causality between two processes measures the level of influence that one process $\{X\}$ has on another process $\{Y\}$. The analysis compares the \textit{error in predicting} the $\{Y\}$ process based on (i) simply the past of $\{Y\}$, and (ii) based on the past of both $\{X\}$ and $\{Y\}$. Many variations of Granger causality have also been considered, including a generalization --- the Directed Information (see~\cite{MasseyCausality,quinn,weissman}) --- an information-theoretic quantity denoted by $I(\m{X}\to\m{Y})$. %So another strategy that could be used to infer causal influence of one process, e.g. $\{X\}$, on another process $\{Y\}$ is to estimate and compare $I(\m{X}\to\m{Y})$ and $I(\m{Y}\to\m{X})$ (normalized by $m$) in the limit of large $m$.

>>>>

This work is in large part motivated by a recent surge of interest in understanding neural circuits -- the connectivity and dynamic activity of different regions of the brain -- and how they give rise to behavior and experience. This is evidenced by the launching of the BRAIN initiative in the US and the Human Brain Project in Europe. To quote from \emph{BRAIN 2025: A Scientific Vision}\footnote{The BRAIN Working Group's report to the Advisory Committee to the Director of the NIH}, we wish to ``map connected neurons in local circuits and distributed brain systems, enabling an understanding of the relationship between neuronal structure and function'', clearly indicating the move towards understanding (a) the connectivity and (b) the computational function of different brain regions. While the question of how the brain computes has been of immense interest for several decades, only recently have measurement techniques become sophisticated enough to be able to simultaneously record the activity of multiple neurons, or multiple neural populations.

In order to understand how the brain performs computations, it could be useful to first understand the directions of \emph{information flow} in various parts of the brain (e.g.~\cite{blinowska2004granger,dhamala2008analyzing,nolte2008robustly,korzeniewska2003determination,schippers2010mapping} etc.). In an effort to make headway on the goals of \emph{BRAIN 2025}, several works use Granger causality (and less often, its information-theoretic generalization -- Directed Information) to understand how this information flows (eg.~\cite{Brovelli2004BetaOscillations,goebel2003investigating,deshpande2008effective}), or to acquire directed maps of functional connectivity (eg.~\cite{friston2013analysing,goebel2003investigating,deshpande2008effective}). For instance, in~\cite{Brovelli2004BetaOscillations}, Granger causal influences that are measured between somatosensory and motor sites are said to ``support the idea that somatosensory feedback provides information to the sensorimotor system that is used to control motor output''. This raises the question: do these directed connectivity maps, as determined by directional causal influence measures such as Granger causality, correctly identify the directions along which information flows in the brain?

% Intro: How GC is used in neuroscience today ---------------------------------

\subsection{How Granger Causality is used in Neuroscience today}
\label{sec:gc-used-in-neuroscience}

Several works have outlined the procedures involved in using Granger Causality to estimate causal influences in the brain (\cite{bernasconi2000bidirectional,Brovelli2004BetaOscillations,Ding2000ShortWindow,Roebroeck2005MappingDirected,Bressler2011WienerGranger,Barnett2014MVGC,Ding2006GrangerCausality}). Here, we briefly describe how Granger Causal influence is quantified, and how it is computed in these works.

Granger causality, as originally described by Granger~\cite{Granger1969InvestigatingCausal}, measures the level of influence that one process $\{X\}$ has on another process $\{Y\}$. The analysis compares the \emph{error in predicting} the $\{Y\}$ process based on (i) simply the past of $\{Y\}$, and (ii) based on the past of both $\{X\}$ and $\{Y\}$. The Granger causality metric is the ratio of these errors, encapsulating the innovations that the process $\{X\}$ causally supplies to the process $\{Y\}$. Many variations of Granger causality have also been considered, including a generalization -- Directed Information (see~\cite{MasseyCausality,quinn,weissman}) -- an information-theoretic quantity denoted by $I(\m{X}\to\m{Y})$. These form possible alternatives for estimating the direction of causal influence, but Directed Information is a generalization of many of these metrics~\cite{quinn}.

The computation of Granger causality, in general, requires that the processes under consideration be stationary. In neuroscientific experiments, this is usually not the case, as the express objective of the experiment is to understand brain function while the brain is subjected to a task. Typically, data analysis techniques either assume stationarity regardless, or circumvent the problem by working with time-windows short enough that the processes \emph{are} in fact stationary within those windows~\cite{Ding2000ShortWindow}. In the latter case, Granger causal influences are computed for each window individually, in order to produce a dynamic description of brain connectivity as the brain carries out the said task. The Granger Causality metrics can be accurately computed even for short time windows, because neuroscientific experiments usually have multiple trials, i.e.\ multiple realizations of the same processes, allowing for fitting regression coefficients accurately~\cite{Ding2000ShortWindow}.

In order to determine the direction of causal influence, the Granger Causality metric (ratio of residual variances) from $\{X\}$ to $\{Y\}$ is \emph{compared} to that from $\{Y\}$ to $\{X\}$. The direction of causal influence is taken to be the direction with a greater Granger Causality metric. Further, this direction of causal influence is interpreted to be the direction of informationf flow -- the interpretation we question in this paper. We also note that many of these works use a spectral version of Granger Causality, that supplies this metric as a function of frequency. It then becomes possible to also determine the brain wave frequency at which these influences occur. However, we restrict our analysis to the simpler non-spectral version of Granger Causality, since it is sufficient to for the purpose of our arguments.

We note here that Granger's original analysis does not compare this metric on forward and reverse links, however, this appears to be commonly done in practice. {\color{red} TODO: Cite Tsachy's paper here.}

@@

% Intro: Previous criticisms of GC --------------------------------------------

\subsection{A short survey of previous criticisms of Granger Causality}
\label{sec:previous-criticisms-of-gc}

@@

\begin{enumerate}
	\item First, Granger Causality suffers from what we call the ``hidden node problem''. If two observed nodes receive causal influences from a third, latent node, then a causal influence may be detected between the observed nodes, even if they are independent of each other~\cite{pearl2009causality-hidden-node}. All nodes need to be observed, therefore, to avoid finding spurious influences.
	\item Second, if the measurements from each node are differentially affected by noise, then the predicted direction of causal influence might be opposite to the true direction (\cite{nalatore2007mitigating,andersson2005testing}). Measurements need to be relatively noiseless and precise in order to obtain the correct direction of causal influece.
	\item Third, subsampling the processes can produce misleading Granger Causal relations~\cite{gong2015discovering}. Performing Granger Causal analysis on subsampled time series can lead one to miss the causal influence. If pre-processing involves subsampling, then this should be done with care.
\end{enumerate}

It is important to note that the technical objections listed above are all concerned with \emph{measurement}. They indicate that incorrect Granger Causal influences may be computed if there is some deficiency in the measurement procedure. These objections can be resolved by taking better measurements (by sampling more nodes, with higher signal-to-noise ratio, etc.).

Our objection, on the other hand, is much more fundamental. \emph{Even if} the measurements are made with infinite accuracy, and the regression coefficients associated with computing Granger Causality are precisely estimated, and the Granger Causality metric is perfectly computed (as is the case in our counter-examples), Granger Causal influence would \emph{still} not give the correct direction of information flow. To our knowledge, this is a completely novel argument against Granger Causality, and one that is far more serious than previous objections, at least as far as its usefulness in understanding brain function is concerned.

% Intro: Our counter-examples -------------------------------------------------

\subsection{Our counter-examples and our main result}
\label{sec:our-counterexamples}

This paper considers two experiments (introduced in Section~\ref{sec:sk}) where a transmitter $Tx$ wants to communicate a message to a receiver $Rx$ in presence of a feedback channel (in one experiment, the feedback link is noiseless, while in the other it is noisy). We assume that the experimenter is able to record the transmissions of $Tx$ and $Rx$ using some probing mechanism. Provided with these measurements, the experimenter wants to estimate the direction of information flow, which in this context is the direction of flow of the message. Our results, derived in Section~\ref{sec:analytical-results} and numerically illustrated in Section~\ref{sec:numerical-results}, show that both Granger causality and Directed Information can incorrectly estimate the direction of information flow. The first experiment considers the (unrealistic) case of communication across noiseless feedback channels. The second experiment allows for noise in the feedback channel. In both cases, linear strategies inspired from the scheme of Schalkwijk and Kailath~\cite{S&K} are used.

Our goal here is to bring out the point that whether Granger causality and Directed Information can be used to interpret direction of information flows is an issue that can be, and perhaps should be, considered using thought experiments on simple communication problems where \emph{information flow direction, and quantity, is already known}. If the direction of causal influence yielded by Granger causality or some other similar measure were to match the known direction of information flow, then that measure can be more confidently used in experiments. While our results strongly suggest that one needs to exercise care in interpreting Granger causality and Directed Information dominance as an indicator for the direction of information flow, there are several shortcomings that need to be addressed in order to understand the issue at depth. These shortcomings are discussed in detail in Section~\ref{sec:objections-and-shortcomings}.

% Intro: Possible objections and shortcomings ---------------------------------

\subsection{Possible objections to, and shortcomings of this work}
\label{sec:objections-and-shortcomings}

A review of a previous submission raised some objections to this work, which we seek to clarify here. Further, our analysis has certain shortcomings, which we acknowledge. These could form the basis for future work.

\begin{enumerate}
	\item Firstly, and most importantly, previous work has already shown that presence of a Granger Causal influence does not imply true causation. Judea Pearl classifies Granger Causality as stemming from a ``statistical'' model, rather than from a ``causal'' model~\cite{pearl2009causality-gc-stat}. A potential issue, therefore, might be the question the novelty of this work, given that Pearl's classification could be seen to clear up the misconception. While our argument is, in spirit, very similar to this criticism of Granger causality, in the situation we study, we ask a different question - we wish to understand whether or not Granger Causality gives the correct direction of \emph{information flow}. Further, we give a concrete counter-example in this context, which is very relevant in the modern neuroscience setting.
	\item It might appear that we make a circular argument while computing the Granger Causal influences in our counter-example, since we supply the model for the stochastic process and use the theoretical coeffiecients of the model directly to compute the Granger Causality metric. However, as we state in Section~\ref{sec:figure-it-out}, we assume that the regression coefficients of the Autoregressive model can be exactly estimated (even if the AR process is non-stationary), and discuss how this might be achieved with the help of multiple trials. {\color{red} (Move to conclusions?)}
	\item Our analysis does not tackle an information source that evolves with time. Hence, our communication process (inspired from the scheme of Schalkwijk and Kailath~\cite{S&K}) is non-ergodic. Since Granger causality is really just relative errors in prediction of a process, in presence or absence of knowledge of another process, we compute the obvious generalization of Granger causality to non-ergodic processes. Nevertheless, future work will address a situation with a linear dynamical system as the information source.
	\item Our experiments restrict themselves to Gaussian noise for simplicity, but neural spiking and spike-rate models for spikes tend to be very different from those used here. This is a clear direction for future work.
	\item The power and energy constraints are somewhat oversimplified to make the analysis simpler. This is for simplicity of exposition. A more general analysis is a trivial extension.
	\item We have also restricted ourselves to analyzing linear feedback communication strategies. In the presence of noise in the feedback link, linear communication strategies are known to be sub-optimal~\cite{YoungHanKimPaper}. In order to make a water-tight argument, we would need to show that Granger Causality fails to correctly predict the direction of information flow, even when an optimal communication strategy -- which would have to be non-linear. This could be scope for future work. However, we do not expect results to change dramatically -- when the feedback link is impaired by noise of only very low variance, the noiseless case should make for a good approximation of the system, and results should degrade gracefully, if there is any degradation at all.
	\item We consider a simple point-to-point network. In general networks, this issue could be even more complex. However, since this issue shows up even for the simplest network, we feel that the problem will only be exacerbated when the network is large.
	\item We would also like to note that this paper is \emph{not} mathematically deep. We believe that this is a strength of our analysis. Even simple counter-examples that do not employ difficult proof techniques are able to show that Granger Causality does not always correctly estimate the direction of information flow. {\color{red} (Move this, but where?)}
\end{enumerate}

\subsection{Discussion}
\begin{enumerate}
	\item \color{red} Some of this misconception might have arisen because of Massey's directed info conservation - we expect that dir info in the forward dirn goes up if the other one goes down - but this is a misnomer. (Do we still need this? I am increasingly of the opinion that directed info / GC capture nothing about the message)
\end{enumerate}

@@

In order to arrive at such an understanding of the brain, a natural method is to probe it to understand its functional connectivity\footnote{This is different from structural or anatomical connectivity in that links may sometimes be present, but not active.}. Analyses are then performed on the time series thus obtained from the probes, in order to say something about the functional connectivity -- which components influence each other, and which direction is this influence in. One commonly used analysis is based on the computation of Granger causality~\cite{Granger} with the measurements from each node. The Granger causality between two processes measures the level of influence that one process $\{X\}$ has on another process $\{Y\}$. The analysis compares the \textit{error in predicting} the $\{Y\}$ process based on (i) simply the past of $\{Y\}$, and (ii) based on the past of both $\{X\}$ and $\{Y\}$. Many variations of Granger causality have also been considered, including a generalization -- the Directed Information (see~\cite{MasseyCausality,quinn,weissman}) -- an information-theoretic quantity denoted by $I(\m{X}\to\m{Y})$. %So another strategy that could be used to infer causal influence of one process, e.g. $\{X\}$, on another process $\{Y\}$ is to estimate and compare $I(\m{X}\to\m{Y})$ and $I(\m{Y}\to\m{X})$ (normalized by $m$) in the limit of large $m$.
